---
title: "Final project ataset: 2024"
output: html_document
date: "2024-05-14"
---











```{r}
rm(list=ls())
library(rjags)
library(bayesplot)
library(jagsUI)
library(loo)
```


# 6. GDP & INFLATION US

The data consists in two time series:

Gross Domestic Product (GDP) 

Consumer Price Index for All Urban Consumers: All Items in U.S. City Average (CPIAUCSL)

(more info in the additional file)

Task. Fit some time series models for the two series (separately). You can to try  AR,MA,GARCH or ARMA. In case you use more model, compare the models 
with some Information criteria (BIC,DIC,WAIC). 
You can also  try a bivariate time series models, e.g. a simple VAR(1) model (in this case, ask to the teacher for more information). 





```{r}
rm(list=ls())

# New Family Houses Sold: United States
# Source: https://fred.stlouisfed.org/series/HSN1F

data = read.csv("data/gdp_inflation.csv",header=T)
data$DATE = as.Date(data$DATE)
gdpData = data$GDP_PC1[1:305]
infData = data$CPIAUCSL_PC1[1:305]
sum(is.na(data$GDP_PC1)) #conta i valori mancanti(NA) in gdp
plot(data$DATE[1:305],gdpData,type="l",xlab="",ylab="GDP",main="GDP+INFL")
lines(data$DATE[1:305],infData,type="l",col="red")
#data$CPIAUCSL_PC1
#data1=as.numeric(data$GDP_PC1[1:305])
```







```{r}
modelARMA.string <-"
model {
  ## parameters: alpha, beta, tau, m0
  # likelihood
  mu[1] <- Y[1]
  Yp[1] <- mu[1]
  eps[1] <- 0
  LogLik[1] <- log(dnorm(Y[1], mu[1], tau))
  
  for (i in 2:N) {
    eps[i] ~ dnorm(0, tau_eps)  # errore bianco
    Y[i] ~ dnorm(mu[i], tau)
    mu[i] <- m0 + alpha * Y[i-1] + beta * eps[i-1]
    Yp[i] ~ dnorm(mu[i], tau) # prediction in sample
    
    # Calcolo della log-verosimiglianza per ogni osservazione
    LogLik[i] <- log(dnorm(Y[i], mu[i], tau))
    
    
  }
  
 
 # prediction out of sample 
  ypOut[1] ~ dnorm(m0 + alpha * Y[N] + beta * eps[N], tau)
  for (k in 2:Npred) {
    eps[N + k - 1] ~ dnorm(0, tau_eps)  # errore bianco per le previsioni future
    ypOut[k] ~ dnorm(m0 + alpha * ypOut[k-1] + beta * eps[N + k - 1], tau)
  }
  sigma2 <- 1 / tau
  sigma_eps2 <- 1 / tau_eps
  # priors 
  alpha ~ dunif(-1.5, 1.5)
  beta ~ dunif(-1.5, 1.5)
  tau ~ dgamma(0.1, 10)
  tau_eps ~ dgamma(0.1, 10)
  m0 ~ dnorm(0.0, 1.0E-4)
}"

```





```{r}

# prepare the data 
gdpData = as.numeric(gdpData)
Ntot=length(gdpData)
Npred=100 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcARMA_GDP <- jags(model.file = textConnection(modelARMA.string),
                           data = line_data,
                           parameters.to.save = c('alpha', 'beta', "sigma2", "sigma_eps2", "m0", "Yp", "ypOut", "LogLik"),
                           n.adapt = 1000, n.iter = 10000, n.chains = 1, n.burnin = 2000)

```


```{r}
t = seq(1, Ntot)
tt = seq(1, N)

# Previsioni in-sample
yp = outputmcmcARMA_GDP$mean$Yp
q1 = outputmcmcARMA_GDP$q2.5$Yp
q2 = outputmcmcARMA_GDP$q97.5$Yp

# Calcolo del valore predittivo medio vero (true predictive mean) per confronto
alpha_hat = outputmcmcARMA_GDP$mean$alpha
beta_hat = outputmcmcARMA_GDP$mean$beta
m0_hat = outputmcmcARMA_GDP$mean$m0



mutrue = m0_hat + alpha_hat * c(NA, gdpData[-length(gdpData)]) + beta_hat * c(0, diff(gdpData))
mutrue = mutrue[-1]
mutrue = c(0, mutrue)
mutrue = mutrue[1:N]

plot(t, gdpData, col="red", ylab="mu_t",
     main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t, gdpData, col="red")
#lines(tt, mutrue[1:N], col="magenta")
lines(tt, mutrue[1:N], col="magenta", type="p")
#points(tt, yp, pch="*", col="blue")
#lines(tt, q1, type="l", col="blue", lwd=1.5)
#lines(tt, q2, type="l", col="blue", lwd=1.5)

# Previsioni out-of-sample
yp_pred = outputmcmcARMA_GDP$mean$ypOut
q1_pred = outputmcmcARMA_GDP$q2.5$ypOut
q2_pred = outputmcmcARMA_GDP$q97.5$ypOut

plot(t, gdpData, col="red", ylab="mu_t", ylim=c(min(q1), max(q2)),
     main="out-of-sample prediction (orange)")
abline(v=N, col="orange")
lines(tt, yp, type="p", pch="*", col="blue")
lines(tt, q1, type="l", col="blue", lwd=1.5)
lines(tt, q2, type="l", col="blue", lwd=1.5)
points(seq((N+1), Ntot, 1), yp_pred, pch="*", col="orange")
lines(seq((N+1), Ntot, 1), q1_pred, col="orange", lwd=1.5)
lines(seq((N+1), Ntot, 1), q2_pred, col="orange", lwd=1.5)
```
```{r}
plot(outputmcmcARMA_GDP$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcARMA_GDP$samples[,c("sigma2")],main="posterior sigma^2")
```

## ARMA(1,1) model Inflation

An ARMA(1) model is a time series model defined as follows. Starting 
from $y_1$ one recursively define for $t \geq 1$
\[
y_{t+1}=\mu+\alpha y_t +\beta\epsilon_t+\epsilon_{t+1} \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]



```{r}

# prepare the data 
infData = as.numeric(infData)
Ntot=length(infData)
Npred=100 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcARMA_INF <- jags(model.file = textConnection(modelARMA.string),
                           data = line_data,
                           parameters.to.save = c('alpha', 'beta', "sigma2", "sigma_eps2", "m0", "Yp", "ypOut"),
                           n.adapt = 1000, n.iter = 10000, n.chains = 1, n.burnin = 2000)

```


```{r}
t = seq(1, Ntot)
tt = seq(1, N)

# Previsioni in-sample
yp = outputmcmcARMA_INF$mean$Yp
q1 = outputmcmcARMA_INF$q2.5$Yp
q2 = outputmcmcARMA_INF$q97.5$Yp

# Calcolo del valore predittivo medio vero (true predictive mean) per confronto
alpha_hat = outputmcmcARMA_INF$mean$alpha
beta_hat = outputmcmcARMA_INF$mean$beta
m0_hat = outputmcmcARMA_INF$mean$m0



mutrue = m0_hat + alpha_hat * c(NA, gdpData[-length(gdpData)]) + beta_hat * c(0, diff(gdpData))
mutrue = mutrue[-1]
mutrue = c(0, mutrue)
mutrue = mutrue[1:N]

plot(t, infData, col="red", ylab="mu_t",
     main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t, infData, col="red")
lines(tt, mutrue[1:N], col="magenta")
lines(tt, mutrue[1:N], col="magenta", type="p")
points(tt, yp, pch="*", col="blue")
lines(tt, q1, type="l", col="blue", lwd=1.5)
lines(tt, q2, type="l", col="blue", lwd=1.5)

# Previsioni out-of-sample
yp_pred = outputmcmcARMA_INF$mean$ypOut
q1_pred = outputmcmcARMA_INF$q2.5$ypOut
q2_pred = outputmcmcARMA_INF$q97.5$ypOut

plot(t, infData, col="red", ylab="mu_t", ylim=c(min(q1), max(q2)),
     main="out-of-sample prediction (orange)")
abline(v=N, col="orange")
lines(tt, yp, type="p", pch="*", col="blue")
lines(tt, q1, type="l", col="blue", lwd=1.5)
lines(tt, q2, type="l", col="blue", lwd=1.5)
points(seq((N+1), Ntot, 1), yp_pred, pch="*", col="orange")
lines(seq((N+1), Ntot, 1), q1_pred, col="orange", lwd=1.5)
lines(seq((N+1), Ntot, 1), q2_pred, col="orange", lwd=1.5)

```
```{r}
plot(outputmcmcARMA_INF$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcARMA_INF$samples[,c("sigma2")],main="posterior sigma^2")
```
```




```




```{r}
# Estrai i parametri stimati dal risultato MCMC
alpha_hat <- outputmcmcARMA_GDP$mean$alpha
beta_hat <- outputmcmcARMA_GDP$mean$beta
tau_hat <- outputmcmcARMA_GDP$meant$tau
m0 <- outputmcmcARMA_GDP$sims.list$m0
params <- c(alpha_hat, beta_hat, tau_hat, m0)

# Verifica che i parametri stimati non siano NA
if (any(is.na(params))) {
  stop("Ci sono parametri stimati che sono NA.")
}

# Verifica che i dati non contengano NA o infiniti
if (any(is.na(gdpData)) || any(is.infinite(gdpData))) {
  stop("I dati contengono NA o valori infiniti.")
}




# Funzione per calcolare la log-verosimiglianza del modello ARMA
logLik_arma <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]
  tau <- params[3]
  m0 <- params[4]
  
  N <- length(data)
  logLik <- 0
  mu <- numeric(N)
  eps <- numeric(N)
  mu[1] <- data[1]
  
  for (i in 2:N) {
    eps[i] <- data[i] - (m0 + alpha * data[i-1] + beta * eps[i-1])
    mu[i] <- m0 + alpha * data[i-1] + beta * eps[i-1]
    
    # Calcola la log-verosimiglianza e verifica eventuali problemi
    logLik_i <- dnorm(data[i], mean=mu[i], sd=sqrt(1/tau), log=TRUE)
    
    if (is.na(logLik_i)) {
      print(paste("NA in logLik a i =", i))
      print(paste("data[i] =", data[i]))
      print(paste("mu[i] =", mu[i]))
      print(paste("m0 =", m0))
      print(paste("alpha =", alpha))
      print(paste("data[i-1] =", data[i-1]))
      print(paste("beta =", beta))
      print(paste("eps[i-1] =", eps[i-1]))
    
      print(paste("tau =", tau))
      stop("Log-verosimiglianza è NA.")
    }
    
    logLik <- logLik + logLik_i
  }
  
  return(logLik)
}


# Calcola la log-verosimiglianza per i dati GDP
logLik_value <- logLik_arma(params, gdpData)

# Numero di parametri stimati
k <- length(params)

# Numero di osservazioni
n <- length(gdpData)

# Calcola il BIC
BIC <- -2 * logLik_value + k * log(n)
BIC


dic_value <- outputmcmcARMA_GDP$DIC
dic_value



loglik_A=outputmcmcARMA_GDP$sims.list$LogLik

#print(loglik_A)

waic_A <- waic(loglik_A)
waic_A

```



```{r}

```


 



```{r}

```




```{r}




```





```{r}

```

 

```{r}

```




```{r}

```








```{r}

```
