---
title: "Final project ataset: 2024"
output: html_document
date: "2024-05-14"
---

```{r}
rm(list=ls())
library(rjags)
library(bayesplot)
library(jagsUI)
```

# 6. GDP & INFLATION US

The data consists in two time series:

Gross Domestic Product (GDP) 

Consumer Price Index for All Urban Consumers: All Items in U.S. City Average (CPIAUCSL)

(more info in the additional file)

Task. Fit some time series models for the two series (separately). You can to try  AR,MA,GARCH or ARMA. In case you use more model, compare the models 
with some Information criteria (BIC,DIC,WAIC). 
You can also  try a bivariate time series models, e.g. a simple VAR(1) model (in this case, ask to the teacher for more information). 



```{r}
rm(list=ls())

# New Family Houses Sold: United States
# Source: https://fred.stlouisfed.org/series/HSN1F

data = read.csv("data/gdp_inflation.csv",header=T)
data$DATE = as.Date(data$DATE)
gdpData = data$GDP_PC1[1:305]
infData = data$CPIAUCSL_PC1[1:305]
sum(is.na(data$GDP_PC1)) #conta i valori mancanti(NA) in gdp
plot(data$DATE[1:305],gdpData,type="l",xlab="",ylab="GDP",main="GDP+INFL")
lines(data$DATE[1:305],infData,type="l",col="red")
#data$CPIAUCSL_PC1
#data1=as.numeric(data$GDP_PC1[1:305])
```

## Moving Average (1)
```{r}
modelMA.string <-"
model {
  ## parameters: alpha, tau, m0
  
  # likelihood 
  for (i in 1:N) {
    e[i] ~ dnorm(0, tau)
  }
  
  mu[1] <- m0 + e[1]
  Yp[1] <- mu[1]
  
  for (i in 2:N) {
    Y[i] ~ dnorm(mu[i], tau)
    mu[i] <- m0 + alpha * e[i-1] + e[i]
    Yp[i] ~ dnorm(mu[i], tau) # prediction in sample
  }
  
  # prediction out of sample 
  e_pred[1] ~ dnorm(0, tau)
  ypOut[1] ~ dnorm(m0 + alpha * e[N] + e_pred[1], tau)
  
  for (k in 2:Npred) {
    e_pred[k] ~ dnorm(0, tau)
    ypOut[k] ~ dnorm(m0 + alpha * e_pred[k-1] + e_pred[k], tau)
  }
  
  sigma2 <- 1 / tau
  
  # priors 
  alpha ~ dunif(-1.5, 1.5)
  tau ~ dgamma(0.1, 0.1) # Adjusted the parameters for a reasonable prior
  m0 ~ dnorm(0.0, 1.0E-4)

}"
```

```{r}
# prepare the data 
gdpData = as.numeric(gdpData)
Ntot=length(gdpData)
Npred=20 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcMA_GDP <- jags(model.file=textConnection(modelMA.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```

```{r}

t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcMA_GDP$mean$Yp
q1=outputmcmcMA_GDP$q2.5$Yp
q2=outputmcmcMA_GDP$q97.5$Yp
#

plot(t,gdpData,col="red",ylab="mu_t",main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t,gdpData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcMA_GDP$mean$ypOut
q1_pred=outputmcmcMA_GDP$q2.5$ypOut
q2_pred=outputmcmcMA_GDP$q97.5$ypOut
#
plot(t,gdpData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
lines(t,gdpData,col="red")
abline(v=N,col="orange")

lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)

```

```{r}
# prepare the data 
infData = as.numeric(infData)
Ntot=length(infData)
Npred=20 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcMA_INF <- jags(model.file=textConnection(modelMA.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```

```{r}

t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcMA_INF$mean$Yp
q1=outputmcmcMA_INF$q2.5$Yp
q2=outputmcmcMA_INF$q97.5$Yp
#

plot(t,infData,col="red",ylab="mu_t",main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t,infData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcMA_INF$mean$ypOut
q1_pred=outputmcmcMA_INF$q2.5$ypOut
q2_pred=outputmcmcMA_INF$q97.5$ypOut
#
plot(t,infData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
lines(t,infData,col="red")
abline(v=N,col="orange")

lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)

```

