---
title: "Final project ataset: 2024"
output: html_document
date: "2024-05-14"
---

```{r}
rm(list=ls())
library(rjags)
library(bayesplot)
library(jagsUI)
```


See also the pdf file with more information on the project. 


# 6. GDP & INFLATION US

The data consists in two time series:

Gross Domestic Product (GDP) 

Consumer Price Index for All Urban Consumers: All Items in U.S. City Average (CPIAUCSL)

(more info in the additional file)

Task. Fit some time series models for the two series (separately). You can to try  AR,MA,GARCH or ARMA. In case you use more model, compare the models 
with some Information criteria (BIC,DIC,WAIC). 
You can also  try a bivariate time series models, e.g. a simple VAR(1) model (in this case, ask to the teacher for more information). 



```{r}
rm(list=ls())

# New Family Houses Sold: United States
# Source: https://fred.stlouisfed.org/series/HSN1F

data = read.csv("data/gdp_inflation.csv",header=T)
data$DATE = as.Date(data$DATE)
gdpData = data$GDP_PC1[1:305]
infData = data$CPIAUCSL_PC1[1:305]
sum(is.na(data$GDP_PC1))
plot(data$DATE[1:305],gdpData,type="l",xlab="",ylab="GDP",main="GDP+INFL")
lines(data$DATE[1:305],infData,type="l",col="red")
#data$CPIAUCSL_PC1
#data1=as.numeric(data$GDP_PC1[1:305])
```

## AR(1) model INF

An AR(1) model is a time series model defined as follows. Starting 
from $y_1$ one recursively define for $t \geq 1$
\[
y_{t+1}=\mu+\alpha y_t +\epsilon_t \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]

```{r}
modelAR.string <-"model {
  ## parameters: alpha,tau,m0
  #likeliohood 
  mu[1]<-Y[1]
  Yp[1]=mu[1]
  for (i in 2:N) {
    Y[i] ~ dnorm(mu[i],tau)
    mu[i]<-m0+alpha*Y[i-1]
    Yp[i] ~ dnorm(mu[i],tau) # prediction in sample
  }
  # prediction out of sample 
  ypOut[1] ~dnorm(m0+alpha*Y[N],tau) 
  for(k in 2:Npred){
  ypOut[k] ~dnorm(m0+alpha*ypOut[k-1],tau) 
  }
  sigma2<-1/tau
  #prior 
  alpha ~ dunif(-1.5,1.5)
  tau ~ dgamma(0.1, 10)
  m0 ~dnorm(0.0, 1.0E-4)
}"
```



```{r}

# prepare the data 
gdpData = as.numeric(gdpData)
Ntot=length(gdpData)-1
Npred=30 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcAR_GDP <- jags(model.file=textConnection(modelAR.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```


```{r}
t=seq(1,Ntot+1)
tt=seq(1,N)
#
yp=outputmcmcAR_GDP$mean$Yp
q1=outputmcmcAR_GDP$q2.5$Yp
q2=outputmcmcAR_GDP$q97.5$Yp
#
alpha_true = 0.9
mtrue = mean(gdpData)
mutrue=alpha_true*gdpData+mtrue
mutrue=c(0,mutrue)
mutrue=mutrue[1:Ntot]
#
plot(t,gdpData,col="red",ylab="mu_t",
     main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t,gdpData,col="red")
lines(tt,mutrue[1:N],col="magenta",type="p")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcAR_GDP$mean$ypOut
q1_pred=outputmcmcAR_GDP$q2.5$ypOut
q2_pred=outputmcmcAR_GDP$q97.5$ypOut
#
plot(t,gdpData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
abline(v=N,col="orange")
lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)
```
```{r}
plot(outputmcmcAR_GDP$samples[,c("alpha")]
,main="posterior alpha")
#plot(outputmcmcAR_GDP$samples[,c("sigma2")],main="posterior sigma^2")
```

## AR(1) model Inflation

An AR(1) model is a time series model defined as follows. Starting 
from $y_1$ one recursively define for $t \geq 1$
\[
y_{t+1}=\mu+\alpha y_t +\epsilon_t \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]



```{r}

# prepare the data 
infData = as.numeric(infData)
Ntot=length(infData)-1
Npred=30 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcAR_INF <- jags(model.file=textConnection(modelAR.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```


```{r}
t=seq(1,Ntot+1)
tt=seq(1,N)
#
yp=outputmcmcAR_INF$mean$Yp
q1=outputmcmcAR_INF$q2.5$Yp
q2=outputmcmcAR_INF$q97.5$Yp
#
alpha_true = 0.9
mtrue = mean(infData)
mutrue=alpha_true*infData+mtrue
mutrue=c(0,mutrue)
mutrue=mutrue[1:Ntot]
#
plot(t,infData,col="red",ylab="mu_t",
     main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t,infData,col="red")
lines(tt,mutrue[1:N],col="magenta",type="p")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcAR_INF$mean$ypOut
q1_pred=outputmcmcAR_INF$q2.5$ypOut
q2_pred=outputmcmcAR_INF$q97.5$ypOut
#
plot(t,gdpData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
abline(v=N,col="orange")
lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)
```
```{r}
plot(outputmcmcAR_INF$samples[,c("alpha")]
,main="posterior alpha")
#plot(outputmcmcAR_INF$samples[,c("sigma2")],main="posterior sigma^2")
```

