---
title: "Final project ataset: 2024"
output: html_document
date: "2024-05-14"
---

```{r}
rm(list=ls())
library(rjags)
library(bayesplot)
library(jagsUI)
library(loo)
```

# 6. GDP & INFLATION US

The data consists in two time series:

Gross Domestic Product (GDP) 

Consumer Price Index for All Urban Consumers: All Items in U.S. City Average (CPIAUCSL)

(more info in the additional file)

Task. Fit some time series models for the two series (separately). You can to try  AR,MA,GARCH or ARMA. In case you use more model, compare the models 
with some Information criteria (BIC,DIC,WAIC). 
You can also  try a bivariate time series models, e.g. a simple VAR(1) model (in this case, ask to the teacher for more information). 

```{r}
# New Family Houses Sold: United States
# Source: https://fred.stlouisfed.org/series/HSN1F

data = read.csv("data/gdp_inflation.csv",header=T)
data$DATE = as.Date(data$DATE)
gdpData = data$GDP_PC1[1:305]
infData = data$CPIAUCSL_PC1[1:305]
sum(is.na(data$GDP_PC1)) #conta i valori mancanti(NA) in gdp
plot(data$DATE[1:305],gdpData,type="l",xlab="",ylab="GDP",main="GDP+INFL")
lines(data$DATE[1:305],infData,type="l",col="red")
#data$CPIAUCSL_PC1
#data1=as.numeric(data$GDP_PC1[1:305])
```

# AR(1) Models

An AR(1) model is a time series model defined as follows. Starting 
from $y_1$ one recursively define for $t \geq 1$
\[
y_{t+1}=\mu+\alpha y_t +\epsilon_t \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]


```{r}
modelAR.string <-"model {
  ## parameters: alpha,tau,m0
  #likeliohood 
  mu[1]<-Y[1]
  Yp[1]=mu[1]
  LogLik[1] = 0
  for (i in 2:N) {
    Y[i] ~ dnorm(mu[i],tau)
    mu[i]<-m0+alpha*Y[i-1]
    Yp[i] ~ dnorm(mu[i],tau) # prediction in sample
    LogLik[i] <- log(dnorm( Y[i],mu[i],tau))
  }
  # prediction out of sample 
  ypOut[1] ~dnorm(m0+alpha*Y[N],tau) 
  for(k in 2:Npred){
  ypOut[k] ~dnorm(m0+alpha*ypOut[k-1],tau) 
  }
  sigma2<-1/tau
  #prior 
  alpha ~ dunif(-1.5,1.5)
  tau ~ dgamma(0.1, 10)
  m0 ~dnorm(0.0, 1.0E-4)
}"
```

```{r}

# prepare the data 
gdpData = as.numeric(gdpData)
Ntot=length(gdpData)
Npred=100 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcAR_GDP <- jags(model.file=textConnection(modelAR.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut","LogLik"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```
## AR(1) model GDP

```{r}
t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcAR_GDP$mean$Yp
q1=outputmcmcAR_GDP$q2.5$Yp
q2=outputmcmcAR_GDP$q97.5$Yp
#

alpha_hatg = outputmcmcAR_GDP$mean$alpha
m0_hatg = outputmcmcAR_GDP$mean$m0
#

plot(t,gdpData,col="red",ylab="mu_t",
     main="data (red), in samp.pred. (blue)")
lines(t,gdpData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcAR_GDP$mean$ypOut
q1_pred=outputmcmcAR_GDP$q2.5$ypOut
q2_pred=outputmcmcAR_GDP$q97.5$ypOut
q11_pred=outputmcmcAR_GDP$q25$ypOut
q22_pred=outputmcmcAR_GDP$q75$ypOut
#
plot(t,gdpData,col="red",ylab="mu_t",ylim=c(min(q1),max(q2)),
     main= "out-of-sample prediction (orange)")
lines(t,gdpData,col="red")
abline(v=N,col="orange")
lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q11_pred,col="orange",lty=2,lwd = 1.5)
lines(seq((N+1),Ntot,1),q22_pred,col="orange",lty=2,lwd = 1.5)
```
```{r}
plot(outputmcmcAR_GDP$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcAR_GDP$samples[,c("sigma2")],main="posterior sigma^2")
```
```{r}
autocorr.plot(outputmcmcAR_GDP$samples[,c("alpha")],main="alpha ACF")
autocorr.plot(outputmcmcAR_GDP$samples[,c("sigma2")],main="sigma2 ACF")
```

## AR(1) model Inflation

```{r}

# prepare the data 
infData = as.numeric(infData)
Ntot=length(infData)
Npred=100 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcAR_INF <- jags(model.file=textConnection(modelAR.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut","LogLik"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```

```{r}
t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcAR_INF$mean$Yp
q1=outputmcmcAR_INF$q2.5$Yp
q2=outputmcmcAR_INF$q97.5$Yp
#

alpha_hati = outputmcmcAR_INF$mean$alpha
m0_hati = outputmcmcAR_INF$mean$m0
#
plot(t,infData,col="red",ylab="mu_t",main="data (red), in samp.pred. (blue)")
lines(t,infData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcAR_INF$mean$ypOut
q1_pred=outputmcmcAR_INF$q2.5$ypOut
q2_pred=outputmcmcAR_INF$q97.5$ypOut
#
plot(t,infData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
lines(t,infData,col="red")
abline(v=N,col="orange")

lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)
```
```{r}
plot(outputmcmcAR_INF$samples[,c("alpha")]
     ,main="posterior alpha")
plot(outputmcmcAR_INF$samples[,c("sigma2")],main="posterior sigma^2")
```


```{r}
autocorr.plot(outputmcmcAR_INF$samples[,c("alpha")],main="alpha ACF")
autocorr.plot(outputmcmcAR_INF$samples[,c("sigma2")],main="sigma2 ACF")
```

# AR(2) model

On the other hand an AR(2) model is a time series model defined as follows. Starting 
from $y_2$ one recursively define for $t \geq 2$
\[
y_{t+1}=\mu+\alpha y_t + \beta y_{t-1}+\epsilon_t  \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]

```{r}
modelAR2.string <-"model {
  ## parameters: alpha,tau,m0
  #likeliohood 
  mu[1]<-Y[1]
  Yp[1]=mu[1]
  mu[2]<-Y[2]
  Yp[2]=mu[2]
  LogLik[1] = 0
  LogLik[2] = 0
  for (i in 3:N) {
    Y[i] ~ dnorm(mu[i],tau)
    mu[i]<-m0+alpha*Y[i-1]+beta*Y[i-2]
    Yp[i] ~ dnorm(mu[i],tau) # prediction in sample
    LogLik[i] <- log(dnorm( Y[i],mu[i],tau))
  }
  # prediction out of sample 
  ypOut[1] ~dnorm(m0+alpha*Y[N]+beta*Y[N-1],tau)
  ypOut[2] ~dnorm(m0+alpha*ypOut[1]+beta*Y[N],tau)
  for(k in 3:Npred){
    ypOut[k] ~dnorm(m0+alpha*ypOut[k-1]+beta*ypOut[k-2],tau) 
  }
  sigma2<-1/tau
  #prior 
  alpha ~ dunif(-1.5,1.5)
  beta ~ dunif(-1.5,1.5)
  tau ~ dgamma(0.1, 10)
  m0 ~dnorm(0.0, 1.0E-4)
}"
```

```{r}

# prepare the data 
Ntot=length(gdpData)
Npred=floor(Ntot/5) # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcAR2_GDP <- jags(model.file=textConnection(modelAR2.string),
                     data=line_data,
                     parameters.to.save= c('alpha','beta',"sigma2","m0","Yp","ypOut", "LogLik"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```


```{r}
t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcAR2_GDP$mean$Yp
q1=outputmcmcAR2_GDP$q2.5$Yp
q2=outputmcmcAR2_GDP$q97.5$Yp
#

alpha_hatg = outputmcmcAR2_GDP$mean$alpha
m0_hatg = outputmcmcAR2_GDP$mean$m0
#

plot(t,gdpData,col="red",ylab="mu_t",
     main="data (red), in samp.pred. (blue)")
lines(t,gdpData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcAR2_GDP$mean$ypOut
q1_pred=outputmcmcAR2_GDP$q2.5$ypOut
q2_pred=outputmcmcAR2_GDP$q97.5$ypOut
q11_pred=outputmcmcAR2_GDP$q25$ypOut
q22_pred=outputmcmcAR2_GDP$q75$ypOut
#
plot(t,gdpData,col="red",ylab="mu_t",ylim=c(min(q1),max(q2)),
     main= "out-of-sample prediction (orange)")
lines(t,gdpData,col="red")
abline(v=N,col="orange")
lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q11_pred,col="orange",lty=2,lwd = 1.5)
lines(seq((N+1),Ntot,1),q22_pred,col="orange",lty=2,lwd = 1.5)
```
```{r}
plot(outputmcmcAR2_GDP$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcAR2_GDP$samples[,c("beta")],main="posterior beta")
plot(outputmcmcAR2_GDP$samples[,c("sigma2")],main="posterior sigma^2")
```
```{r}
autocorr.plot(outputmcmcAR2_GDP$samples[,c("alpha")],main="alpha ACF")
autocorr.plot(outputmcmcAR2_GDP$samples[,c("beta")],main="beta ACF")
autocorr.plot(outputmcmcAR2_GDP$samples[,c("sigma2")],main="sigma2 ACF")
```

## AR(2) model Inflation

An AR(1) model is a time series model defined as follows. Starting 
from $y_1$ one recursively define for $t \geq 1$
\[
y_{t+1}=\mu+\alpha y_t +\epsilon_t \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]

```{r}

# prepare the data 
Ntot=length(infData)
Npred=floor(Ntot/5) # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcAR2_INF <- jags(model.file=textConnection(modelAR2.string),
                     data=line_data,
                     parameters.to.save= c('alpha', 'beta',"sigma2","m0","Yp","ypOut","LogLik"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```


```{r}
t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcAR2_INF$mean$Yp
q1=outputmcmcAR2_INF$q2.5$Yp
q2=outputmcmcAR2_INF$q97.5$Yp
#

alpha_hati = outputmcmcAR2_INF$mean$alpha
m0_hati = outputmcmcAR2_INF$mean$m0
#
plot(t,infData,col="red",ylab="mu_t",main="data (red), in samp.pred. (blue)")
lines(t,infData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcAR2_INF$mean$ypOut
q1_pred=outputmcmcAR2_INF$q2.5$ypOut
q2_pred=outputmcmcAR2_INF$q97.5$ypOut
#
plot(t,infData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
lines(t,infData,col="red")
abline(v=N,col="orange")

lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)
```
```{r}
plot(outputmcmcAR2_INF$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcAR2_INF$samples[,c("beta")],main="posterior beta")
plot(outputmcmcAR2_INF$samples[,c("sigma2")],main="posterior sigma^2")
```


```{r}
autocorr.plot(outputmcmcAR2_INF$samples[,c("alpha")],main="alpha ACF")
autocorr.plot(outputmcmcAR2_INF$samples[,c("beta")],main="beta ACF")
autocorr.plot(outputmcmcAR2_INF$samples[,c("sigma2")],main="sigma2 ACF")
```

# Model Selection

```{r}
# Funzione per calcolare la log-verosimiglianza del modello ARMA
logLik_ar <- function(params, data) {
  alpha <- params[1]
  #beta <- params[2]
  tau <- params[2]
  m0 <- params[3]
  
  N <- length(data)
  logLik <- 0
  mu <- numeric(N)
  eps <- numeric(N)
  mu[1] <- data[1]
  
  for (i in 2:N) {
    eps[i] <- data[i] - (m0 + alpha * data[i-1] + eps[i-1])
    mu[i] <- m0 + alpha * data[i-1] + eps[i-1]
    
    # Calcola la log-verosimiglianza e verifica eventuali problemi
    logLik_i <- dnorm(data[i], mean=mu[i], sd=sqrt(1/tau), log=TRUE)
    
    if (is.na(logLik_i)) {
      print(paste("NA in logLik a i =", i))
      print(paste("data[i] =", data[i]))
      print(paste("mu[i] =", mu[i]))
      print(paste("m0 =", m0))
      print(paste("alpha =", alpha))
      print(paste("data[i-1] =", data[i-1]))
      #print(paste("beta =", beta))
      print(paste("eps[i-1] =", eps[i-1]))
    
      print(paste("tau =", tau))
      stop("Log-verosimiglianza è NA.")
    }
    
    logLik <- logLik + logLik_i
  }
  
  return(logLik)
}

# Funzione per calcolare la log-verosimiglianza del modello ARMA
logLik_ar2 <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]
  tau <- params[3]
  m0 <- params[4]
  
  N <- length(data)
  logLik <- 0
  mu <- numeric(N)
  eps <- numeric(N)
  mu[1] <- data[1]
  
  for (i in 3:N) {
    eps[i] <- data[i] - (m0 + alpha * data[i-1] +beta * data[i-2] + eps[i-1])
    mu[i] <- m0 + alpha * data[i-1] +beta * data[i-2]+ eps[i-1]
    
    # Calcola la log-verosimiglianza e verifica eventuali problemi
    logLik_i <- dnorm(data[i], mean=mu[i], sd=sqrt(1/tau), log=TRUE)
    
    if (is.na(logLik_i)) {
      print(paste("NA in logLik a i =", i))
      print(paste("data[i] =", data[i]))
      print(paste("mu[i] =", mu[i]))
      print(paste("m0 =", m0))
      print(paste("alpha =", alpha))
      print(paste("data[i-1] =", data[i-1]))
      print(paste("beta =", beta))
      print(paste("eps[i-1] =", eps[i-1]))
    
      print(paste("tau =", tau))
      stop("Log-verosimiglianza è NA.")
    }
    
    logLik <- logLik + logLik_i
  }
  
  return(logLik)
}

### BIC GDP AR(1)
alpha_hat <- outputmcmcAR_GDP$mean$alpha
tau_hat <- outputmcmcAR_GDP$mean$tau
m0_hat <- outputmcmcAR_GDP$sims.list$m0
params <- c(alpha_hat, tau_hat, m0_hat)

# Calcola la log-verosimiglianza per i dati GDP
logLik_valueAR_GDP <- logLik_ar(params, gdpData)

# Numero di parametri stimati
k <- length(params)

# Numero di osservazioni
n <- length(gdpData)

# Calcola il BIC
BIC_AR_GDP <- -2 * logLik_valueAR_GDP + k * log(n)

### BIC Inflation AR(1)
alpha_hat <- outputmcmcAR_INF$mean$alpha
tau_hat <- outputmcmcAR_INF$mean$tau
m0_hat <- outputmcmcAR_INF$sims.list$m0
params <- c(alpha_hat, tau_hat, m0_hat)

# Calcola la log-verosimiglianza per i dati GDP
logLik_valueAR_INF <- logLik_ar(params, gdpData)

# Numero di parametri stimati
k <- length(params)

# Numero di osservazioni
n <- length(gdpData)

# Calcola il BIC
BIC_AR_INF <- -2 * logLik_valueAR_INF + k * log(n)

### BIC GDP AR(2)
alpha_hat <- outputmcmcAR2_GDP$mean$alpha
beta_hat <- outputmcmcAR2_GDP$mean$beta
tau_hat <- outputmcmcAR_GDP$mean$tau
m0_hat <- outputmcmcAR_GDP$sims.list$m0
params <- c(alpha_hat, beta_hat, tau_hat, m0_hat)

# Calcola la log-verosimiglianza per i dati GDP
logLik_valueAR2_GDP <- logLik_ar2(params, gdpData)

# Numero di parametri stimati
k <- length(params)

# Numero di osservazioni
n <- length(gdpData)

# Calcola il BIC
BIC_AR2_GDP <- -2 * logLik_valueAR2_GDP + k * log(n)

### BIC Inflation AR(1)
alpha_hat <- outputmcmcAR2_INF$mean$alpha
beta_hat <- outputmcmcAR2_INF$mean$beta
tau_hat <- outputmcmcAR2_INF$mean$tau
m0_hat <- outputmcmcAR2_INF$sims.list$m0
params <- c(alpha_hat, beta_hat, tau_hat, m0_hat)

# Calcola la log-verosimiglianza per i dati GDP
logLik_valueAR2_INF <- logLik_ar2(params, gdpData)

# Numero di parametri stimati
k <- length(params)

# Numero di osservazioni
n <- length(gdpData)

# Calcola il BIC
BIC_AR2_INF <- -2 * logLik_valueAR2_INF + k * log(n)
```

```{r}
### Calcolo del WAIC per GDP AR(1)
loglik=outputmcmcAR_GDP$sims.list$LogLik # model without change point
print(length(loglik))
GDPAR1_waic<- waic(loglik)
print(GDPAR1_waic)
```
```{r}
### Calcolo del WAIC per CPI AR(1)
loglik=outputmcmcAR_INF$sims.list$LogLik # model without change point
print(length(loglik))
INFAR1_waic<- waic(loglik)
print(INFAR1_waic)
```

```{r}
### Calcolo del WAIC per GDP AR(2)
loglik<- outputmcmcAR2_GDP$sims.list$LogLik # model without change point
print(length(loglik))
sum(is.na(loglik))
GDPAR2_waic<- waic(loglik)
print(GDPAR2_waic)
```
```{r}
### Calcolo del WAIC per CPI AR(2)
loglik=outputmcmcAR2_INF$sims.list$LogLik # model without change point
print(length(loglik))

INFAR2_waic<- waic(loglik)
print(INFAR2_waic)
```

```{r}
BIC_AR_GDP
BIC_AR2_GDP
```

```{r}
BIC_AR_INF
BIC_AR2_INF
```


```{r}
dic_ARGDP <- outputmcmcAR_GDP$DIC
dic_ARINF <- outputmcmcAR_INF$DIC
dic_AR2GDP <- outputmcmcAR2_GDP$DIC
dic_AR2INF <- outputmcmcAR2_GDP$DIC
```

```{r}
dic_ARGDP
dic_AR2GDP
```

```{r}
dic_ARINF
dic_AR2INF
```

According to both, the results of the BIC and the DIC for the proposed models, it is reasonable to state that the AR(1) model fits best the data available since it overfits it less, while maintaining more or less the same effectivity on the prediction.

# ARMA Models   

```{r}
modelARMA.string <-"
model {
  ## parameters: alpha, beta, tau, m0
  # likelihood
  mu[1] <- Y[1]
  Yp[1] <- mu[1]
  eps[1] <- 0
  LogLik[1] <- log(dnorm(Y[1], mu[1], tau))
  
  for (i in 2:N) {
    eps[i] ~ dnorm(0, tau_eps)  # errore bianco
    Y[i] ~ dnorm(mu[i], tau)
    mu[i] <- m0 + alpha * Y[i-1] + beta * eps[i-1]
    Yp[i] ~ dnorm(mu[i], tau) # prediction in sample
    
    # Calcolo della log-verosimiglianza per ogni osservazione
    LogLik[i] <- log(dnorm(Y[i], mu[i], tau))
    
    
  }
  
 
 # prediction out of sample 
  ypOut[1] ~ dnorm(m0 + alpha * Y[N] + beta * eps[N], tau)
  for (k in 2:Npred) {
    eps[N + k - 1] ~ dnorm(0, tau_eps)  # errore bianco per le previsioni future
    ypOut[k] ~ dnorm(m0 + alpha * ypOut[k-1] + beta * eps[N + k - 1], tau)
  }
  sigma2 <- 1 / tau
  sigma_eps2 <- 1 / tau_eps
  # priors 
  alpha ~ dunif(-1.5, 1.5)
  beta ~ dunif(-1.5, 1.5)
  tau ~ dgamma(0.1, 10)
  tau_eps ~ dgamma(0.1, 10)
  m0 ~ dnorm(0.0, 1.0E-4)
}"

```

## ARMA(1,1) model GDP

An ARMA(1) model is a time series model defined as follows. Starting 
from $y_1$ one recursively define for $t \geq 1$
\[
y_{t+1}=\mu+\alpha y_t +\beta\epsilon_t+\epsilon_{t+1} \qquad 
\epsilon_t \stackrel{iid}{\sim} \mathcal{N}(0,\sigma^2)
\]

```{r}

# prepare the data 
gdpData = as.numeric(gdpData)
Ntot=length(gdpData)
Npred=100 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcARMA_GDP <- jags(model.file = textConnection(modelARMA.string),
                           data = line_data,
                           parameters.to.save = c('alpha', 'beta', "sigma2", "sigma_eps2", "m0", "Yp", "ypOut", "LogLik"),
                           n.adapt = 1000, n.iter = 10000, n.chains = 1, n.burnin = 2000)

```


```{r}
t = seq(1, Ntot)
tt = seq(1, N)

# Previsioni in-sample
yp = outputmcmcARMA_GDP$mean$Yp
q1 = outputmcmcARMA_GDP$q2.5$Yp
q2 = outputmcmcARMA_GDP$q97.5$Yp

# Calcolo del valore predittivo medio vero (true predictive mean) per confronto
alpha_hat = outputmcmcARMA_GDP$mean$alpha
beta_hat = outputmcmcARMA_GDP$mean$beta
m0_hat = outputmcmcARMA_GDP$mean$m0



mutrue = m0_hat + alpha_hat * c(NA, gdpData[-length(gdpData)]) + beta_hat * c(0, diff(gdpData))
mutrue = mutrue[-1]
mutrue = c(0, mutrue)
mutrue = mutrue[1:N]

plot(t, gdpData, col="red", ylab="mu_t",
     main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t, gdpData, col="red")
#lines(tt, mutrue[1:N], col="magenta")
lines(tt, mutrue[1:N], col="magenta", type="p")
#points(tt, yp, pch="*", col="blue")
#lines(tt, q1, type="l", col="blue", lwd=1.5)
#lines(tt, q2, type="l", col="blue", lwd=1.5)

# Previsioni out-of-sample
yp_pred = outputmcmcARMA_GDP$mean$ypOut
q1_pred = outputmcmcARMA_GDP$q2.5$ypOut
q2_pred = outputmcmcARMA_GDP$q97.5$ypOut

plot(t, gdpData, col="red", ylab="mu_t", ylim=c(min(q1), max(q2)),
     main="out-of-sample prediction (orange)")
abline(v=N, col="orange")
lines(tt, yp, type="p", pch="*", col="blue")
lines(tt, q1, type="l", col="blue", lwd=1.5)
lines(tt, q2, type="l", col="blue", lwd=1.5)
points(seq((N+1), Ntot, 1), yp_pred, pch="*", col="orange")
lines(seq((N+1), Ntot, 1), q1_pred, col="orange", lwd=1.5)
lines(seq((N+1), Ntot, 1), q2_pred, col="orange", lwd=1.5)
```
```{r}
plot(outputmcmcARMA_GDP$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcARMA_GDP$samples[,c("sigma2")],main="posterior sigma^2")
```

## ARMA(1,1) model Inflation


```{r}

# prepare the data 
infData = as.numeric(infData)
Ntot=length(infData)
Npred=100 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcARMA_INF <- jags(model.file = textConnection(modelARMA.string),
                           data = line_data,
                           parameters.to.save = c('alpha', 'beta', "sigma2", "sigma_eps2", "m0", "Yp", "ypOut"),
                           n.adapt = 1000, n.iter = 10000, n.chains = 1, n.burnin = 2000)

```


```{r}
t = seq(1, Ntot)
tt = seq(1, N)

# Previsioni in-sample
yp = outputmcmcARMA_INF$mean$Yp
q1 = outputmcmcARMA_INF$q2.5$Yp
q2 = outputmcmcARMA_INF$q97.5$Yp

# Calcolo del valore predittivo medio vero (true predictive mean) per confronto
alpha_hat = outputmcmcARMA_INF$mean$alpha
beta_hat = outputmcmcARMA_INF$mean$beta
m0_hat = outputmcmcARMA_INF$mean$m0



mutrue = m0_hat + alpha_hat * c(NA, gdpData[-length(gdpData)]) + beta_hat * c(0, diff(gdpData))
mutrue = mutrue[-1]
mutrue = c(0, mutrue)
mutrue = mutrue[1:N]

plot(t, infData, col="red", ylab="mu_t",
     main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t, infData, col="red")
lines(tt, mutrue[1:N], col="magenta")
lines(tt, mutrue[1:N], col="magenta", type="p")
points(tt, yp, pch="*", col="blue")
lines(tt, q1, type="l", col="blue", lwd=1.5)
lines(tt, q2, type="l", col="blue", lwd=1.5)

# Previsioni out-of-sample
yp_pred = outputmcmcARMA_INF$mean$ypOut
q1_pred = outputmcmcARMA_INF$q2.5$ypOut
q2_pred = outputmcmcARMA_INF$q97.5$ypOut

plot(t, infData, col="red", ylab="mu_t", ylim=c(min(q1), max(q2)),
     main="out-of-sample prediction (orange)")
abline(v=N, col="orange")
lines(tt, yp, type="p", pch="*", col="blue")
lines(tt, q1, type="l", col="blue", lwd=1.5)
lines(tt, q2, type="l", col="blue", lwd=1.5)
points(seq((N+1), Ntot, 1), yp_pred, pch="*", col="orange")
lines(seq((N+1), Ntot, 1), q1_pred, col="orange", lwd=1.5)
lines(seq((N+1), Ntot, 1), q2_pred, col="orange", lwd=1.5)

```
```{r}
plot(outputmcmcARMA_INF$samples[,c("alpha")],main="posterior alpha")
plot(outputmcmcARMA_INF$samples[,c("sigma2")],main="posterior sigma^2")
```


```{r}
# Estrai i parametri stimati dal risultato MCMC
alpha_hat <- outputmcmcARMA_GDP$mean$alpha
beta_hat <- outputmcmcARMA_GDP$mean$beta
tau_hat <- outputmcmcARMA_GDP$meant$tau
m0 <- outputmcmcARMA_GDP$sims.list$m0
params <- c(alpha_hat, beta_hat, tau_hat, m0)

# Verifica che i parametri stimati non siano NA
if (any(is.na(params))) {
  stop("Ci sono parametri stimati che sono NA.")
}

# Verifica che i dati non contengano NA o infiniti
if (any(is.na(gdpData)) || any(is.infinite(gdpData))) {
  stop("I dati contengono NA o valori infiniti.")
}




# Funzione per calcolare la log-verosimiglianza del modello ARMA
logLik_arma <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]
  tau <- params[3]
  m0 <- params[4]
  
  N <- length(data)
  logLik <- 0
  mu <- numeric(N)
  eps <- numeric(N)
  mu[1] <- data[1]
  
  for (i in 2:N) {
    eps[i] <- data[i] - (m0 + alpha * data[i-1] + beta * eps[i-1])
    mu[i] <- m0 + alpha * data[i-1] + beta * eps[i-1]
    
    # Calcola la log-verosimiglianza e verifica eventuali problemi
    logLik_i <- dnorm(data[i], mean=mu[i], sd=sqrt(1/tau), log=TRUE)
    
    if (is.na(logLik_i)) {
      print(paste("NA in logLik a i =", i))
      print(paste("data[i] =", data[i]))
      print(paste("mu[i] =", mu[i]))
      print(paste("m0 =", m0))
      print(paste("alpha =", alpha))
      print(paste("data[i-1] =", data[i-1]))
      print(paste("beta =", beta))
      print(paste("eps[i-1] =", eps[i-1]))
    
      print(paste("tau =", tau))
      stop("Log-verosimiglianza è NA.")
    }
    
    logLik <- logLik + logLik_i
  }
  
  return(logLik)
}


# Calcola la log-verosimiglianza per i dati GDP
logLik_value <- logLik_arma(params, gdpData)

# Numero di parametri stimati
k <- length(params)

# Numero di osservazioni
n <- length(gdpData)

# Calcola il BIC
BIC <- -2 * logLik_value + k * log(n)
BIC


dic_value <- outputmcmcARMA_GDP$DIC
dic_value



loglik_A=outputmcmcARMA_GDP$sims.list$LogLik

#print(loglik_A)

waic_A <- waic(loglik_A)
waic_A

```



# Moving Average (1)
```{r}
modelMA.string <-"
model {
  ## parameters: alpha, tau, m0
  
  # likelihood 
  for (i in 1:N) {
    e[i] ~ dnorm(0, tau)
  }
  
  mu[1] <- m0 + e[1]
  Yp[1] <- mu[1]
  
  for (i in 2:N) {
    Y[i] ~ dnorm(mu[i], tau)
    mu[i] <- m0 + alpha * e[i-1] + e[i]
    Yp[i] ~ dnorm(mu[i], tau) # prediction in sample
  }
  
  # prediction out of sample 
  e_pred[1] ~ dnorm(0, tau)
  ypOut[1] ~ dnorm(m0 + alpha * e[N] + e_pred[1], tau)
  
  for (k in 2:Npred) {
    e_pred[k] ~ dnorm(0, tau)
    ypOut[k] ~ dnorm(m0 + alpha * e_pred[k-1] + e_pred[k], tau)
  }
  
  sigma2 <- 1 / tau
  
  # priors 
  alpha ~ dunif(-1.5, 1.5)
  tau ~ dgamma(0.1, 0.1) # Adjusted the parameters for a reasonable prior
  m0 ~ dnorm(0.0, 1.0E-4)

}"
```

```{r}
# prepare the data 
gdpData = as.numeric(gdpData)
Ntot=length(gdpData)
Npred=20 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=gdpData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcMA_GDP <- jags(model.file=textConnection(modelMA.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```

```{r}

t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcMA_GDP$mean$Yp
q1=outputmcmcMA_GDP$q2.5$Yp
q2=outputmcmcMA_GDP$q97.5$Yp
#

plot(t,gdpData,col="red",ylab="mu_t",main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t,gdpData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcMA_GDP$mean$ypOut
q1_pred=outputmcmcMA_GDP$q2.5$ypOut
q2_pred=outputmcmcMA_GDP$q97.5$ypOut
#
plot(t,gdpData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
lines(t,gdpData,col="red")
abline(v=N,col="orange")

lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)

```

```{r}
# prepare the data 
infData = as.numeric(infData)
Ntot=length(infData)
Npred=20 # horizon for out-of-sample prediction
N=Ntot-Npred
data_subsample=infData[1:N]

line_data <- list("Y" =data_subsample,"N" = length(data_subsample),"Npred"=Npred)

outputmcmcMA_INF <- jags(model.file=textConnection(modelMA.string),
                     data=line_data,
                     parameters.to.save= c('alpha',"sigma2","m0","Yp","ypOut"), 
                     n.adapt=1000, n.iter=10000,n.chains = 1,n.burnin = 2000)

```

```{r}

t=seq(1,Ntot)
tt=seq(1,N)
#
yp=outputmcmcMA_INF$mean$Yp
q1=outputmcmcMA_INF$q2.5$Yp
q2=outputmcmcMA_INF$q97.5$Yp
#

plot(t,infData,col="red",ylab="mu_t",main="data (red), true pred. means (magenta), in samp.pred. (blue)")
lines(t,infData,col="red")
points(tt,yp,pch="*",col="blue")
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
#
yp_pred=outputmcmcMA_INF$mean$ypOut
q1_pred=outputmcmcMA_INF$q2.5$ypOut
q2_pred=outputmcmcMA_INF$q97.5$ypOut
#
plot(t,infData,col="red",ylab="mu_t",ylim=c(min(q1_pred),max(q2_pred)),
     main= "out-of-sample prediction (orange)")
lines(t,infData,col="red")
abline(v=N,col="orange")

lines(tt,yp,type="p",pch="*",col="blue",)
lines(tt,q1,type="l",col="blue",lwd = 1.5)
lines(tt,q2,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred,col="orange",lwd = 1.5)

```



# VAR(1) model INF

```{r}
modelVAR.string <- "model {
   ## parameters:  a11, a12, a21, a22, [m01 m02]
   
   #likelihood
   mu[1:2,1]<- Y[1:2,1]
   Yp[1:2,1]<- mu[1:2,1]
   for (i in 2:N) {
      mu[1,i]<- m0[1] + (a11*Y[1,i-1]) + (a12*Y[2, i-1])
      mu[2,i]<- m0[2] + (a21*Y[1, i-1]) +(a22*Y[2, i-1])
      Y[1:2,i] ~ dmnorm(mu[1:2,i], omega[1:2,1:2])
      Yp[1:2,i] ~ dmnorm(mu[1:2,i], omega)
   }
   #prediction out of sample
   mp[1,1]<- m0[1] + (a11* Y[1,N]) + (a12*Y[2,N])
   mp[2,1]<- m0[2] + (a21*Y[2,N]) + (a22*Y[2,N])
   ypOut[1:2,1] ~ dmnorm(mp[1:2,1], omega)
   for ( k in 2:Npred){
      mp[1,k]<- m0[1] +(a11*ypOut[1,k-1]) + (a12 * ypOut[2,k-1])
      mp[2,k]<- m0[2] +(a21*ypOut[1,k-1]) + (a22*ypOut[2,k-1])
      ypOut[1:2,k] ~ dmnorm(mp[1:2,k], omega)
   }
   # priors
   a11 ~ dunif(-1.5, 1.5)
   a12 ~ dunif(-1.5, 1.5)
   a21 ~ dunif(-1.5, 1.5)
   a22 ~ dunif(-1.5, 1.5)
   m<- 3
   omega ~ dwish(R,m)
   m0 ~ dmnorm(vec, S)

}"
```


```{r}
gdpData = as.numeric(gdpData)
infData = as.numeric(infData)
data = rbind(gdpData, infData)
Ntot = length(gdpData)-1
Npred=30 # horizon for out-of-sample prediction
N=Ntot-Npred
data_sub_gdp=gdpData[1:N]
data_sub_inf=infData[1:N]
data_subsample= rbind(data_sub_gdp, data_sub_inf)
print(data_subsample[1:2, 1:5])
print(dim(data_subsample))

R <- matrix(0, nrow = 2, ncol = 2)
R[1,1]<- 5
R[2,2]<- 5

S <- matrix(0, nrow = 2, ncol = 2)
S[1,1]<-0.2
S[2,2]<-0.2
vec = c(0,0)
print(N)

line_data <- list("Y" = data_subsample, "N" = N, "Npred"=Npred, "R"= R, "S" = S, "vec"= vec)

outputVAR <- jags(model.file=textConnection(modelVAR.string),
                  data = line_data,
                  parameters.to.save = c("a11", "a12", "a21", "a22", "m0","Yp", "ypOut"),
                  n.adapt = 1000, n.iter = 10000, n.chains = 1, n.burnin = 2000
)

```


```{r}
t=seq(1,Ntot+1)
tt=seq(1,N)
yp_gdp = outputVAR$mean$Yp[1,]
yp_inf = outputVAR$mean$Yp[2,]
q1_gdp = outputVAR$q2.5$Yp[1,]
q2_gdp = outputVAR$q97.5$Yp[1,]
q1_inf = outputVAR$q2.5$Yp[2,]
q2_inf = outputVAR$q97.5$Yp[2,]

plot(t, gdpData, col="red",ylab = "annual growth rate", main = " gdp data (red), gdp sample prediction (blue)" )
lines(t, gdpData, col = "red")
points(tt, yp_gdp, pch = "*", col = "blue")
lines(tt, q1_gdp, type = "l", col = "blue", lwd = 1.5)
lines(tt, q2_gdp, type = "l", col = "blue", lwd = 1.5)

yp_gdp_pred = outputVAR$mean$ypOut[1,]
q1_pred_gdp =outputVAR$q2.5$ypOut[1,]
q2_pred_gdp =outputVAR$q97.5$ypOut[1,]
yp_inf_pred = outputVAR$mean$ypOut[2,]
q1_pred_inf = outputVAR$q2.5$ypOut[2,]
q2_pred_inf = outputVAR$q97.5$ypOut[2,]

plot(t,gdpData,col="red",ylab="annual growth rate",ylim=c(min(q1_pred_gdp),max(q2_pred_gdp)),
     main= "out-of-sample prediction (orange)")
abline(v=N,col="orange")
lines(tt,yp_gdp,type="p",pch="*",col="blue")
lines(tt,q1_gdp,type="l",col="blue",lwd = 1.5)
lines(tt,q2_gdp,type="l",col="blue",lwd = 1.5)
points(seq((N+1),Ntot,1),pch="*",yp_gdp_pred,col="orange")
lines(seq((N+1),Ntot,1),q1_pred_gdp,col="orange",lwd = 1.5)
lines(seq((N+1),Ntot,1),q2_pred_gdp,col="orange",lwd = 1.5)

plot(t, infData, col = "deeppink", ylab = "annual growth rate of Inf",main = "inf data (magenta), inf sample pred (purple)")
lines(t, infData, col = "deeppink")
points(tt, yp_inf, pch = "*", col = "springgreen")
lines(tt, q1_inf, type = "l", col = "springgreen", lwd = 1.5)
lines(tt, q2_inf, type = "l", col = "springgreen", lwd = 1.5)

plot(t, infData, col = "deeppink", ylab = "annual growth rate", ylim = c(min(q1_pred_inf), max(q2_pred_inf)),
     main = "out of sample prediction (orange)" )
abline(v = N, col = "orange")
lines (tt, yp_inf, type = "p", pch = "*", col = "springgreen")
lines(tt, q1_inf, type = "l", col = "springgreen", lwd = 1.5)
lines(tt, q2_inf, type = "l", col = "springgreen", lwd = 1.5)
points(seq((N+1), Ntot, 1), pch = "*", yp_inf_pred, col = "orange" )
lines(seq((N+1), Ntot,1), q1_pred_inf, col = "orange", lwd = 1.5)
lines(seq((N+1), Ntot,1), q2_pred_inf, col = "orange", lwd = 1.5)

#plot(outputVAR)

```
